Synthesising Concise Termination and Non-Termination Proofs for Bit-Vector Programs

### 2014-07-08 papers/termination/paper.tex
Edits

\begin{theorem}
 \label{thm:generalised-sat}
 The generalised termination formula for loop $L$ (Definiton~\ref{def:general-termination}) is satisfiable
 when $P_N$ and $P_T$ range over \mbox{$\mathcal{L}$-computable} functions.
\end{theorem}

@@

\begin{proof}
 By Theorem~\ref{thm:generalised-sat}, the specification \lstinline!spec! is satisfiable.  Corollary~\ref{cor:cegis-newc}
 shows that Algorithm~\ref{alg:cegis} is semi-complete, and so the call to {\sc Synthesise} terminates with a witness $(P_T, P_N)$.
 If $L$ terminates then $P_T$ is a termination proof (Theorem~\ref{thm:ct}).  Otherwise $P_N$ is a non-termination proof
 (Theorem~\ref{thm:snt}).  Exactly one of these cases holds, and the background theory being decidable means that each
 call to \lstinline|decide()| will terminate, which allows us to determine which of $P_T$ and $P_N$ is a correct proof.
\end{proof}

@@

Theorem~\ref{thm:synth-semicomplete} provides a time bound -- the CEGIS loop can iterate at most
$2^bk$ times where $k$ is the number of variables used by $L$ and
$b$ is the number of bits used to store each variable.
The $i$th iteration of the loop requires solving 2 NP-complete problems of size $i \times l \times \| L\|$
with $l$ the length of the program to be synthesised, and $\| L \| $ the size of the program-under-analysis
computing $ L $.  So if
the CEGIS loop iterates $n$ times, the
runtime is bounded by $2 \times n \times NP(n \times l \times \| L \|)$, which is bounded
by $2^{bk + 1} NP(2^{bk} \times l \times \| L \|)$.  Clearly the runtime is affected most strongly by $n$,
the total number of iterations of the CEGIS loop.

@@

Let us consider the number of iterations of the CEGIS loop $n$ required for a specification
$\sigma$.  Since we enumerate candidate programs in order of length, we are always synthesising
programs with length no greater than $K(\sigma)$ (since when we enumerate the first correct program,
we will terminate).  So it is clear that the space of solutions we search over is the space
of functions computed by $\mathcal{L}$-programs of length no greater than $K(\sigma)$.  Let's
denote this set $\mathcal{L}(K(\sigma))$.  Clearly, $\| \mathcal{L}(K(\sigma)) \| \leq O(2^{K(\sigma)})$
since there are $O(2^{K(\sigma)}$ \emph{programs} of length $K(\sigma)$ and some functions
will be computed by more than one program.

Each iteration of the CEGIS loop distinguishes at least one incorrect function from the set of correct
functions, so we can see that the loop will iterate no more than $\| \mathcal{L}(K(\sigma)) \|$ times.
Therefore another bound on our runtime is:
$$O\left (2^{K(\sigma)} \times NP \left ( K(\sigma) \times \| L \| \right) \right) = O \left(2^{K(\sigma)} \right)$$
Which means that for even moderately large $K(\sigma)$ our procedure is intractably slow.  Conversely
if $K(\sigma)$ is small we will be able to find a termination proof, regardless of other parameters
such as the size of the program under analysis.  To put this another way, under the assumption
that $K(\sigma)$ is small for a program's termination problem, we will find a termination proof
quite rapidly.

>>>>

\begin{theorem}
 \label{thm:generalised-sat}
 The generalised termination formula {\bf [GT]} for any loop $L$ is satisfiable
 when $P_N$ and $P_T$ range over \mbox{$\mathcal{L}$-computable} functions.
\end{theorem}

@@

\begin{proof}
 By Theorem~\ref{thm:generalised-sat}, the specification \lstinline!spec! is satisfiable.  Corollary~\ref{cor:cegis-newc}
 shows that Algorithm~\ref{alg:cegis} is semi-complete, and so the call to {\sc Synthesise} terminates with a witness $(P_T, P_N)$.
 If $L$ terminates then $P_T$ is a termination proof (Theorem~\ref{thm:ct}).  Otherwise $P_N$ is a non-termination proof
 (Theorem~\ref{thm:snt}).  Exactly one of these cases holds, and the background theory \newC being decidable means that each
 call to \lstinline|decide()| will terminate, which allows us to determine which of $P_T$ and $P_N$ is a correct proof.
\end{proof}

@@

Theorem~\ref{thm:synth-semicomplete} provides a time bound -- the CEGIS loop can iterate at most
$2^{bk}$ times where $k$ is the number of variables used by $L$ and
$b$ is the number of bits used to store each variable.
The $i$th iteration of the loop requires solving 2 NP-complete problems of size $i \times l \times \| L\|$
with $l$ the length of the program to be synthesised, and $\| L \| $ the size of the program-under-analysis
computing $ L $.  So if
the CEGIS loop iterates $n$ times, the
runtime is bounded by
$$2 \times n \times NP(n \times l \times \| L \|) \leq 2^{bk + 1} NP(2^{bk} \times l \times \| L \|)$$
Clearly the runtime is affected most strongly by $n$, the total number of iterations of the CEGIS loop.

@@

Let us consider the number of iterations of the CEGIS loop $n$ required for a specification
$\sigma$.  Since we enumerate candidate programs in order of length, we are always synthesising
programs with length no greater than $K(\sigma)$ (since when we enumerate the first correct program,
we will terminate).  So the space of solutions we search over is the space
of functions computed by $\mathcal{L}$-programs of length no greater than $K(\sigma)$.  Let's
denote this set $\mathcal{L}(K(\sigma))$.
Since there are $O(2^{K(\sigma)}$ \emph{programs} of length $K(\sigma)$ and some functions
will be computed by more than one program, we have $\| \mathcal{L}(K(\sigma)) \| \leq O(2^{K(\sigma)})$.

Each iteration of the CEGIS loop distinguishes at least one incorrect function from the set of correct
functions, so the loop will iterate no more than $\| \mathcal{L}(K(\sigma)) \|$ times.
Therefore another bound on our runtime is:
$$O\left (2^{K(\sigma)} \times NP \left ( K(\sigma) \times \| L \| \right) \right) = O \left(2^{K(\sigma)} \right)$$
So for even moderately large $K(\sigma)$ our procedure is intractably slow.  Conversely
if $K(\sigma)$ is small we will be able to find a termination proof, regardless of other parameters
such as the size of the program under analysis.  To put this another way, under the assumption
that $K(\sigma)$ is small for a program's termination problem, we will find a termination proof
quite rapidly.
